\providecommand{\main}{../../..}
\documentclass[\main/thesis.tex]{subfiles}

\begin{document}

\section{Mathematical Literature Review}
There exists an extensive amount of literature that study cancer initiation \cite{Durrett, Gentry, Paterson}, progression \cite{Beerenwinkel, Enderling}, metastasis \cite{Franssen}, treatment (chemotherapy, immunotherapy, radiation) \cite{dePillis, Radunskaya, Enderling}, and effects of various micro-environmental and external factors on cancer development \cite{Gerlee} from the perspective of mathematical analysis. However, the only mathematical analysis on field cancerization that could be found at the time of writing, were spatial stochastic models in \textcite{Foo} and \textcite{Ryser}.

\subsection{Review of Foo et al.}
\textcite{Foo} describe field cancerization as the process of primary tumours forming from genetically altered fields of premalignant cells that have high chances of progression to malignancy. Also they state that the premalignant fields can cause recurrent tumours if not excised with the primary tumour during surgery. The main objectives of their study were to:
\begin{enumerate}
  \item develop a spatial evolutionary framework for field cancerization;
  \item describe the size and geometry of the premalignant fields at the moment 
        of tumour initiation;
  \item determine the risk of multi-focal lesions, recurrence timing, and clonal 
        origin of recurrent tumours;
  \item discover the effects of different characteristics of tissue and cancer 
        type on 2 and 3.
\end{enumerate}

The domain of their model is a lattice of dimension ($d$) 1, 2, or 3 wherein each lattice point is occupied by one cell. The dimension is generally either 1 or 2 because although epithelial tissues are 3D it is sufficient to consider just 1 or 2D approximations of the areas of interest. For example due to the ratio of tube radius to the length of mammary ducts of the breast, renal tubules of the kidney and bronchi tubes of the lung can be approximated with $d{=}1$. Also cancer initiation in the squamous epithelium of the cervix, bladder, and oral cavity can be considered a $d {=} 2$ process as it occurs in the basal layer which is only 1-2 cells thick. In the lattice model, each cell reproduces by placing its offspring randomly in one of the $2d$ neighbours at the rate equal to their fitness, $s {\ge} 0$, at exponential waiting time. A cell having a fitness advantage has increased reproductive rate or avoidance of apoptotic signals. The cell type is determined by the fitness, $s$, which they also describe as the number of genetic hits a cell has accumulated. A cell of type-0 have fitness normalized to 1, $s {=} 1$, and are considered to be wild-type or normal. The type-0 cell can become a type-1 cell by acquiring the first type of mutation at the rate $u_1$. In general a type-$i$ cell has a fitness advantage of $1 + s_i$ relative to the type-(i-1) and is assumed to have acquired all the mutations up to mutation $i$. If it acquires the $(i+1)-th$ mutation at rate $u_{i{+}1}$ then it will become a type-(i+1) cell. The mutation rates are quite small, $u_i << 1$, and so the domain has to be large to increase the chance a cell will acquire a mutation. The previously stated definitions and properties make up what is called the spatial Moran model. 

The lattice is initialized with cells having type-0. The starting time is at the end of tissue development and start of the tissue renewal phase. Since this time is difficult to acquire it is challenging to find the time of cancer initiation, $\sigma_k$. It is assumed that when a cell has developed $k$ mutations and so is of type-$k$ that cancer has initialized and the simulation is stopped. The critical number of mutations $k$ is set based upon the type of cancer because different cancers have different mutation thresholds. It is important to note that the model assumes all $k$-mutations positively influence cancer initialization and have the effect of increased cell growth and/or reduction in apoptotic signalling. Another assumption they used was, that all their mutations come from random mutations, which occur very rarely and are generally fixed by the bodies DNA repair process, before they have an effect. \textcite{Foo} ignore the genetic hits that cause selective disadvantage because the result is that the cell and its' progeny die too quickly. Cell death and reproduction dynamics are accomplished similarly to the biased voter model \cite{Liggett}. 

To determine the probability of a mutant clone population of type-1 surviving they considered a one mutation model with cells of type-0 having fitness 1 and type-1 having fitness 1+s. The lattice is initialized with all type-0 cells except the cell at $x{=}0$, this allows the lattice to be either finite or infinite. In the finite case a Williams-Bjerknes model \cite{Williams} is used, otherwise a biased voter model is used \cite{Liggett}. The number of type-1 cells is a jump process in which a discrete time random walk moves one up with probability 
$$
\frac{s}{1{+}s}
$$ 
or one down with probability 
$$
\frac{1}{1{+}s},
$$ 
since the only possible events are type-0 which is replaced by type-1 (jump up) and type-1 which is replaced by a type-0 (jump down). Thus the probability that a mutant clone population of type-1 survives is 
$$
\frac{s}{1{+}s} \approx s
$$
(when $s << 1$). A mutant clone population with fitness $s$ is successful if it reaches size $>> \frac{1}{s}$ as this results in a negligible chance of its extinction. Unsuccessful type-1 mutant clone populations typically have a space-time volume of order:
$$
l(s) {=} \begin{cases}
            s^{-2}, & d{=}1 \\
            s^{-1}\ln(s^{-1}), & d{=}2 \\
            s^{-1}, & d{=} 3 
          \end{cases},
$$
which represents the magnitude of the domain a mutant clone population will occupy. The probability of a mutant clone population of type-1 surviving still holds for spatial Moran models as long as 
$$
\frac{1}{u_1} >> l(s)^{\frac{d{+}2}{2}},
$$
where $u_1$ is the mutation rate for type-0 cells to become type-1 cells.As a result of the previous condition if the number of type-1 cells is significantly less than the total number of cells, $N$, for all time, then successful type-1 mutations arrive as a Poisson process at rate 
$$
\frac{Nu_1s}{1{+}s}.
$$
Upon considering a mesoscopic model (a hybrid model that includes elements from both the macroscopic and microscopic scales) they allow type-1 mutations arrive as Poisson with rate $Nu_1$, distributed uniformly at random in the domain. Each mutation event has two potential outcomes:
\begin{itemize}
  \item[(a)] With probability $\frac{s}{1{+}s}$ the mutation is successful and 
             the clonal expansion is approximated with a ball whose radius grows 
             deterministically with macroscopic growth rate $c_d(s)$.
  \item[(b)] With probability $\frac{1}{1{+}s}$ the mutation is unsuccessful and 
             the clone is conditioned to go extinct.
\end{itemize}

Having acquired the basic probabilistic and size aspects of the model they moved to a two mutation model so that it could be fit in the context of field cancerization whereby type-0 cells have fitness 1, type-1 cells are premalignant with fitness $1 {+} s_1$ relative to type-0, and type-2 are cancer cells with fitness $1 {+} s_2$ relative to type-1. If $s_1{=}s_2{=}s > 0$ the timing of cancer initiation is controlled by the limiting value of the meta-parameter: 
$$\Gamma {=} (Nu_1s)^{d{+}1}(c_d^d(s)u_2s)^{-1},$$ 
which is the number of type-1 clones needed to generate a successful type-2 mutation. If $\Gamma \rightarrow 0$ then the first successful type 2 mutation will occur in the first successful type 1 clone. Varying the meta-parameter $\Gamma$ results in the following three regimes:
\begin{itemize}
  \item[(a)] Regime 1 (R1): When $\Gamma {<} 1$ the first successful type-2 
             mutation occurs within the expanding clone of the first successful 
             type-1 mutation. The time at cancer initiation, $\sigma_2$, is 
             exponential and does not depend on the spatial dimension.
  \item[(b)] Regime 2 (R2): For $\Gamma {\in} (10, 100)$ the first successful 
             type-2 mutation occurs within one of the several successful type-1 
             clones. The time at cancer initiation, $\sigma_2$, is not 
             exponential and depends on the spatial dimension.
  \item[(c)] Regime 3 (R3): When $\Gamma {>} 1000$ the first successful type-2 
             mutation occurs after many successful type-1 mutations. It can 
             arise from a successful or unsuccessful type-1 mutation. The time 
             at cancer initiation, $\sigma_2$, is a mixture distribution of 
             these events.
\end{itemize}
The borderline regimes $R1/R2$ and $R2/R3$ occur for $\Gamma {\in} [1, 10]$ and \newline$\Gamma {\in} [100, 1000]$, respectively. For all further analysis \textcite{Foo} assumed that in $R3$ successful type-2 mutation arise only from successful type-1. As the number of cells increases, the time that the first type-2 mutation arises, $\sigma_2$, decreases, and thus cancer initiation occurs earlier. When $\sigma_2$ is small, there is a diffuse premalignant field and a large number of independent lesions. However, when $\sigma_2$ is large there is a single premalignant field that contains the initial tumour cell. Assuming that the time of diagnosis, $T_D$, is independent of $\sigma_2$ then the premalignant field at time of diagnosis, $\sigma_2 {+} T_D$, can be characterized by the field at $\sigma_2$ together with the distribution of $T_D$. 

The size of an initiating clone population is picked from a distribution generated around the sizes of the clones at the time the initiated mutation arose. If the premalignant field is left during surgery, they found that tumours appearing later have a higher recurrence probability. As $u_1$ increases, the premalignant field is made up of an increasing number of independent type-1 patches and the model moves towards R2 and R3. As the number of type-1 patches increases, so does the chance of a type-2 cell being formed and hence $\sigma_2$ (time of first type-2 cell forming) decreases. As $u_2$ increases, the model moves towards R1 in which fewer type-1 clones are required to produce the first successful type-2 cell, and the size of the type-1 field decreases. In R1 and R2 the total distant field size is of the same order of magnitude as the local field size, whereas in R3 it is significantly larger. Thus, secondary tumour recurrences for cancer types in R3 are more likely to come from the distant field and as a result are not directly related to the primary tumour because the distant fields will be genetically distinct. In R1 the expected number of small cell groups peaks and then declines as larger cell groups begin to dominate, whereas in R2 and R3 small and large cell groups coexist for a longer period of time. A larger premalignant field increases the chance of fast recurrence. In R1 local recurrence is more likely, however when $\sigma_2$ is large, then it is slightly more likely to come back in the distant fields. In R2 and R3 the overall probability of local and distant recurrences is comparable, however when $\sigma_2$ is small, recurrence is more likely to occur in distant fields and when $\sigma_2$ is larger local recurrence is more likely. 

\textcite{Foo} also found the distribution of the local field radius, distribution of the area of the local field at $\sigma_2$, size-distribution of the distant field clones at $\sigma_2$, distribution of the number of field patches, probability of a second field tumour having formed before a given time, and probability that the distant field at the time of initiation gives rise to a second primary tumour at some point in time. 

A limitation of \textcite{Foo} is that they use a specific sequence of mutations, although it is well known that mutations are not sequential and they occur in random order. They assume the micro-environment is static and uniform, which brings about differences in the timing and intensity of malignancy. They use circular growth for the cell groups which does not allow them to grow in every direction or shapes other than circular. They assume a two-hit mutation model, when in reality it is dependent on the cancer as to how many mutations are required. There is no genetic heterogeneity within the clone groups. They don't consider carcinogens as mutagens but instead consider only random mutations, which would greatly dwindle the process of field cancerization since carcinogenic onslaught is considered a significant mutagen. The phenotypic actions they consider are proliferation and apoptosis, but they do not include differentiation and quiescence.  They ignore tissue structures, thus boundaries and growth limitations will not occur. Finally, they include only normal tissue cells, where as, most tissue also includes blood cells, fat cells, and stem cells. 

\subsection{Review of Ryser et al.}
\textcite{Ryser} applied the model described in \textcite{Foo} to head and neck squamous cell carcinoma (HNSCC) which is commonly a result of field cancerization. They specifically investigated HPV-negative cases, patients that had a history of smoking, and age of diagnosis. \textcite{Ryser} attribute the difficulty of detecting the premalignant fields surrounding the primary tumour in clinical practise to a lack of understanding of the dynamics and geometry of the fields. Since HNSCC are formed in stratified epithelia, the growth and renewal of the tissue is accomplished by stem cells (SC), which they call progenitor cells (PC). The PC are located in the basal layer of the tissue, and they renew the tissue by producing transit amplifying cells (TAC), which have limited proliferative potential. These TAC move upwards through the tissue generating new cells and within a few weeks or less are sloughed off. As a result of the short lifespan of the TAC, they are not good candidates for becoming mutated and thus do not create neoplastic lesions. Therefore \textcite{Ryser} only consider the PC in their microscopic model. Cells mutate from normal to cancerous through accumulation of genetic aberrations. When a normal PC acquires growth advantage its' progeny start spreading through the epithelium. Since the number and timing of genetic alterations changes from patient to patient \textcite{Ryser} decided to look at phenotypic progression instead of genotypic progression. 

The three histopathological stages of epithelial dysplasia (precancerous stages) are mild, moderate and severe (carcinoma in situ [CIS]). Based on these three stages \textcite{Ryser} consider the following four type of cells: normal cells (type 0), mildly dysplastic cells (type 0*), moderately dysplastic cells (type 1), and severely dysplastic cells (type 2). They use the stochastic Moran model on a regular two-dimensional lattice as described in \textcite{Foo}. \textcite{Ryser} initialize the model with normal PCs (type 0) with proliferative rate $f_0$. A type 0 cell becomes a type 0* cell at the rate $u_{1, a}$, type 0* becomes type 1 at the rate $u_{1, b}$, and type 1 becomes type 2 at the rate $u_2$. Biologically the proliferative rate of type 0 and type 0* is the same so type 0* has proliferative rate $f_0$. Based on the previous result they computed the probability of a type 0* cell becoming a type 1 cell, thus resulting in a rate $v_{01}$ that a type 0 cell becomes a type 1 cell. Type 1 cells have a proliferative advantage over type 0 and type 0* cells due to having a fitness advantage $s_1$, and so its' proliferative rate is given by $f_1 = f_0(1 + s_1)$. Similarly the type 2 cells have a proliferative rate of $f_2 = f_1(1 + s_2)$. 

\textcite{Ryser} model the time between onset of carcinoma in situ (CIS) and diagnosis using an exponentially distributed random variable with rate $\Psi$. To analytically compute the waiting times and field geometries they use the mesoscopic approximation to the spatial model from \textcite{Foo}. In this model the arrival of expanding type 1 clones is a stochastic Poisson process with rate $Nu_1\frac{s_1}{1+s_1}$, where $N$ is the total number of cells. The factor $\overline{s_1}=\frac{s_1}{1+s_1}$ represents the idea that progeny of a new type 1 cell either go to extinction with probability $1-\overline{s_1}$ or expand indefinitely with probability $\overline{s_1}$. 

According to a theorem, expanding type 1 clones asymptotically grow as a convex symmetric shape with constant radial growth rate $c_2$. In \textcite{Foo} it was found in particular that the convex symmetric shape can be approximated as a disk or circle. The rate $c_2$ depends on the selective advantage $s_1$. For small $s_1$ it scales as $c_s(s_1) \sim \sqrt{4\pi s_1 / \log{(1/s_1)}}$, where $f(s) \sim g(s)$ means that $f(s)/g(s) \rightarrow 1$ as $s \rightarrow 0$. 

For larger values of $s_1$ the relationship had to be numerically computed, in particular for $s_1 > 0.5$, \textcite{Ryser} found an approximately linear dependence given by $c_2(s_1) \approx 0.6s_1 + 0.22$. \textcite{Ryser} allow the existence of multiple precancer fields of type 1 cells within the model. 

To estimate and compute the parameters for their model they use age-specific incidence rates from the Surveillance, Epidemiology, and End Results (SEER) program of the National Cancer Institute (18 registries, 2000-2012) in a Bayesian framework. Since \textcite{Ryser} only considered HPV-negative cancers they restricted to only HNSCC within the lip, tongue, floor of the mouth, gum and other mouth, hypopharynx, and larynx. They computed the number of susceptible individuals and the number of cancer cases diagnosed for the following age groups: 15-19, 20-24, ..., 80-84, and 85+. The final reduction to the data used was to consider only smokers within each considered age groups, as tobacco consumption is a major cause of HPV-negative HNSCC. 

\textcite{Ryser} computed the survival function, the probability density function of the local field radius, and the probability of harboring at least two clonally unrelated fields in the head and neck region with respect to the mean age at smoking initiation to diagnosis with invasive cancer. They found that there is a strong dependence of the local field size on age at diagnosis, with a doubling of the expected field diameter between ages at diagnosis of 50 and 90 years. Further the probability of harboring multiple clonally unrelated fields at the time of diagnosis were found to increase substantially with patient age. As a result of these discoveries \textcite{Ryser} suggest that patient age at diagnosis is a critical predictor of the size and multiplicity of precancerous lesions. 

\subsection{Review of Gerlee et al.}
The next set of literature discusses mathematical techniques that were used for the model that will be proposed in this thesis. The first paper by \textcite{Gerlee} inspired the general framework of the mathematical model. \textcite{Gerlee} created a hybrid cellular automaton to model the effect of various micro-environmental factors on solid tumour growth. Their model is a hybrid cellular automaton because the rule of the automaton depends upon the output of a neural network and partial differential equations. The cellular automaton comprises of two cell types; an empty cell (normal cell) and a tumour cell. It is initialized by setting all the automaton elements to empty except the middle four cells which are occupied by tumour cells. The neural network is used to approximate the relationship between the micro-environmental variables and the phenotype of a cell. The partial differential equations are used to model the spread of the various micro-environmental variables in the domain of consideration. 

For the neural network they use a multi-layer perceptron (MLP), with input being the output of the partial differential equation for the cell at a location $(x, y)$ and output being a vector of likelihoods of a phenotype and movement occurring at a time-step. The hidden layer of the MLP represents the genes and hence the neural network attempts to replicate the genotype-phenotype relationship. They consider the phenotypes proliferation (P), quiescence (Q), and apoptosis (A). Each time-step represents a cell cycle so that a single phenotypic action will occur for each cell. The maximum of the likelihoods between P, Q, and A determines which phenotypic action occurs. If the likelihood of movement is sufficiently large then the cell is allowed to move. 

Each of the actions also has some restrictions based upon the cells \newline
metabolism and adhesion but these will be ignored due to both of these aspects of the cell not being modelled in this thesis. All of the partial differential equations are chemical field equations of the form: 
$$
\frac{\partial c(x, y, t)}{\partial t} = D \Delta c(x, y, t) \pm f(x, y, t).
$$
Where $c(x, y, t)$ is the concentration of the micro-environmental variable, $D$ is the diffusion coefficient, $\Delta$ is the Laplacian operator, and 
$$f(x, y, t) = 
\begin{cases} 
  0, &\text{If the automaton element at } (x, y) \text{ is empty} \\ 
  rF(x, y), &\text{If the automaton element is occupied} 
\end{cases};$$
where $r$ is the base consumption/production rates and $F(x, y)$ is the modulated energy consumption of the individual cell occupying the automaton element at $(x, y)$. The hybrid cellular automaton goes through the following process for each cell and time-step:
\begin{enumerate}
  \item The input to the neural network is sampled from the local environment.
  \item The MLP is computed to obtain the phenotype and likelihood of movement.
  \item The cell consumes nutrients according to the chosen phenotype 
        and the metabolic pathway is chosen.
  \item The chosen phenotype is carried out.
  \item If movement is activated and the cell has not divided it tries to move to 
        a neighbouring cell location.
\end{enumerate}

This thesis aims to answer the following questions around field cancerization:
\begin{enumerate}
    \item What degree of carcinogenic onslaught is necessary for field cancerization to occur? Which carcinogens are the most aggressive?
    \item How long before a cancer field is formed? How long before the first tumour cell within the field is formed?
    \item Is the field formed via monoclonal origin, polyclonal origin, or a mixture of both? Which type of origin is the most common?
    \item How long does it take for a tumour to be large enough such that it is detectable by physicians? Once a tumour is detected, what size is the surrounding field?
    \item How long does it take a recurrence to occur after removal of the field versus the field remaining intact?
\end{enumerate}

\end{document}